{\LARGE Basis Choice for Scalar-on-Function Regression \\ with Applications to Near-Infrared Spectroscopy}

### 2022-02-09 Paper/main.tex
Small Equation correction
Small correction
small corrections

	In the following, we will call the $\hat{\nu}^m(s)$ functional principal components to distinguish the sample analogs from the theoretical Eigenfunctions $\nu^m(s)$. As in the case of ordinary principal components, the number of functional principal components corresponding to non-zero Eigenvalues is limited (cf. chapter 8.2.3 \cite{ramsay_functional_2005}). As each curve is infinite-dimensional, there is no upper limit to this number due to the dimensionality. However, the number of curves still imposes an upper limit of $N-1$ non-zero Eigenvalues, where $N$ is the number of curves in the data set.
	These sample analogs naturally lead to the following representation of each sample curve $x_i(t)$.
	\begin{equation}
		x_i(t) = \hat{\mu}(t) + \sum_{j = 1}^{N-1} \hat{\xi}_{i}^{m} \hat{\nu}^{m}(t)
	\end{equation}
	where the $\hat{\xi}_{i}^m$ are derived as 
	\begin{equation}
		\hat{\xi}_i^m(\omega) = \langle x_i - \hat{\mu}, \hat{\nu}^m\rangle = \int_{0}^{1} \left(x_i(s) - \hat{\mu}(s)\right) \hat{\nu}^m(s) \mathrm{d}s
	\end{equation}
	In reality, these calculations are often implemented using basis representations of both the functional principal components $\hat{\nu}^m$ and the observations $x_i(t)$ leading to the following representation. For the sake of clarity, the following equation assumes that the bases used for the expansion of both the observations and the coefficient function are proper bases of $\mathbb{L}^2[0,1]$ and can therefore be used to express the corresponding objects exactly.
	\begin{equation}\label{FPCA_basis_expansion}
		\begin{split}
			\hat{\xi}_{i}^m & = \int_{0}^{1} {\color{red}\left( x_i(s) - \hat{\mu}(s)\right)} {\color{blue}\hat{\nu}^m(s)} \mathrm{d}s
			= \int_{0}^{1} {\color{red}\left(\sum_{j \in \mathcal{I}} a_{i,j} \phi_j(s)\right)} {\color{blue}\left(\sum_{k \in \mathcal{L}} d_{k}^m \psi_{k}(s)\right)} \mathrm{d}s \\
			& = \int_{0}^{1} {\color{red}\left(\sum_{j = 1}^{J} a_{i,j} \phi_j(s) + \delta_i^J(s)\right)} {\color{blue}\left(\sum_{k = 1}^{K} d_{k}^m \psi_{k}(s) + \delta_{\nu}^{m,K}(s)\right)} \mathrm{d}s \\
			& = \sum_{j = 1}^{J} \left[a_{i,j}\sum_{k = 1}^{K} d_{k}^m \int_{0}^{1} \phi_j(s) \psi_{k}(s)\mathrm{d}s \right] +  \sum_{k = 1}^{K} d_{k}^m \int_{0}^{1} \delta_i^J(s) \psi_{k}(s) \mathrm{d}s \\
			& \quad + \sum_{j = 1}^{J} a_{i,j} \int_{0}^{1}\phi_j(s) \delta_{\nu}^{m,K}(s) \mathrm{d}s + \int_{0}^{1} \delta_i^J(s) \delta_{\nu}^{m,K}(s) \mathrm{d}s
		\end{split}
	\end{equation}
	In practice, a typical choice is to use the same basis $\left(\phi_j(t)\right)_{j \in \mathcal{I}}$ and the same truncation parameter $L$ for the basis expansion of the demeaned observations $\left(x_i(t) - \hat{\mu}(t)\right)$ and the functional principal components $\hat{\nu}^m$. This leads to the following simplification of Equation \ref{FPCA_basis_expansion}.
	\begin{equation}\label{score_approx}
		\begin{split}
			\hat{\xi}_{i}^m &= \sum_{j = 1}^{L} \left[a_{i,j}\sum_{k = 1}^{L} d_{k}^m \int_{0}^{1} \phi_j(s) \phi_{k}(s)\mathrm{d}s \right] +  \sum_{k = 1}^{L} d_{k}^m \int_{0}^{1} \delta_i^L(s) \phi_{k}(s) \mathrm{d}s \\
			& \quad + \sum_{j = 1}^{L} a_{i,j} \int_{0}^{1}\phi_j(s) \delta_{\beta}^L(s) \mathrm{d}s + \int_{0}^{1} \delta_i^L(s) \delta_{\beta}^L(s) \mathrm{d}s
		\end{split}
	\end{equation}
	And we can define the following objects:
		\begin{align}
			\tilde{\xi}^{m,L}_{i} & := \sum_{j = 1}^{L} \left[a_{i,j}\sum_{k = 1}^{L} b_{k}^m \int_{0}^{1} \phi_j(s) \phi_{k}(s)\mathrm{d}s \right] 
			& \delta_{\xi, i}^L & := \hat{\xi}_{i}^m - \tilde{\xi}^{m,L}_{i} \\
			\tilde{\nu}^{m,L}(t) & := \sum_{k = 1}^{L} d_{k}^m \phi_{k}(t) 
			& \delta_{\nu}^{m, L}(t) & := \hat{\nu}^m(t) - \tilde{\nu}^{m,L}(t)
		\end{align}

@@

	To exploit the regularity of the curves of the spectra, we introduced different basis functions in Section \ref{bases} and demonstrated the importance of the truncation parameter $K$ for the estimation in Section \ref{Scalar_on_function_regression}. For the simulation study, we rely on the introduced estimation strategies with the introduced basis functions and focus on choosing the truncation parameter $L$ as well as the number of FPC's, which is possibly affected by $L$, by ten-fold cross-validation using the prediction mean-squared error. The number of FPC's is in practice chosen by using the lowest number of FPC's that explains a specified amount of variability (cf. \cite{kokoszka_introduction_2017}), which might not be optimal since FPC's with smaller Eigenvalues may have a more significant influence on the prediction (cf. \cite{Jolliffe_1982}). This might apply to this simulation too since certain Eigenfunctions could correspond to certain chemical combinations and overtones in the absorption bands of the spectrum that could have high predictive power, but explain only little variability in the NIR curves shown in Appendix \ref{NIR}.\\

@@

		 Two different error-terms $\epsilon$ were created by first generating an i.i.d. standard normal error term and then multiplying it by two error variations $\sigma_e $. The error variations represent different signal-to-noise ratios of the responses to test the methods with low and high amounts of noise in the responses. They are created such that the squared multiple correlation coefficient $R^2 = var(\langle X, f\rangle) / (var(\langle X, f\rangle) + \sigma^2_{e})$ is equal to 0.9 and 0.6. The two error terms are then used to generate two sets of responses for $f \in \{f_1(t), f_2(t)\}$	
		\begin{equation}
			\begin{split}
				Y_{1,f} & = \langle NIR, f\rangle + Z  \biggl\lbrack\frac{var(\langle NIR_{sim}, f\rangle)}{0.9} - var(\langle NIR_{sim}, f\rangle)\biggr\rbrack \\
				Y_{2,f} & = \langle NIR_{sim}, f\rangle + Z  \biggl\lbrack\frac{var(\langle NIR_{sim}, f\rangle)}{0.6} - var(\langle NIR_{sim}, f\rangle)\biggr\rbrack
			\end{split}
		\end{equation}

@@

	\subsubsection{Basis Expansion Regression}
	The following results originate from the \hyperref[basis_exp_transf]{Estimation using Basis-Representation}, in which we transform the smooth curves to perform scalar-on-function regression. Examining the results, it appears that the cross-validated MSPE exhibits a convex behavior in the number of basis functions.

	\paragraph{Monomial Basis}
	Due to the increasing collinearity problems for higher numbers of Monomial basis functions, simulations were conducted up until the sixth monomial, which already shows signs of numerical instability. For reasons outlined in \ref{Monomial_basis}, they seem suited for $f_1$, where it shows a better performance than B-splines for this coefficient function. A hypothesis for this could be that $f_1$ is an entire function, which can be well approximated with a power series. In the case of $f_2$, this basis shows the weakest performance out of all basis, for which the plots in Figures \ref{basis_expansion_2_1} and \ref{basis_expansion_2_2} provide evidence. It seems like $\hat{\beta}(t)$ is not changing in the amount of noise and shows exaggerated behavior at the boundaries. This weakness is especially pronounced in the MSPE for $f_2,Y_1$. For $f_1$, the simulation selects 5(3) and for $f_2$ 5(5) Monomial basis functions for the high(low) signal-to-noise ratio. 
	\vspace{-0.2cm}

	\paragraph{B-spline Basis}
	Simulations with B-spline basis functions were possible from 4 to 18 since from 18 onwards, the simulations were running into problems of numerical instability.
	Function $f_1$ requires 5(4) B-spline basis functions for the high(low) signal-to-noise ratio to obtain the best fit for the B-spline basis and shows the worst performance of the three bases. An explanation might be its overripe behavior at the boundaries and the exaggeration of the peculiarities of $f_1$ (Figures \ref{basis_expansion_1_1} and \ref{basis_expansion_1_2}), which is especially pronounced for the lower signal-to-noise ratio. For $f_2$, 11(6) B-spline basis functions are chosen for the high(low) signal-to-noise ratio. The B-spline basis outperforms the Monomial basis in $f_2$ but comes second to the Fourier basis. While the basis seems to recognize the peculiarities in $f_2,Y_1$, it seems to struggle for the noisy responses in $f_2,Y_2$ (Figures \ref{basis_expansion_2_1} and \ref{basis_expansion_2_2}). With the low signal-to-noise ratio in the responses, the simulation chooses a smaller number of basis functions, which could prevent fitting this scalar noise using $\hat{\beta}(t)$.
	\vspace{-0.2cm}

@@

		Therefore, we will rerun all specifications for the gasoline data set and use the results of the simulation study to improve our understanding of the results. The application is designed analogously to the simulation study: The 60 spectra of the gasoline data set will be used to estimate a coefficient function, predict the reported octane numbers, and evaluate the results via prediction MSE using 12 fold cross-validation with 5 elements per fold. In total, we conducted 1000 repetitions for each setting.

>>>>

	In the following, we will call the $\hat{\nu}^m(s)$ functional principal components to distinguish the sample analogs from the theoretical Eigenfunctions $\nu^m(s)$. As in the case of ordinary principal components, the number of functional principal components corresponding to non-zero Eigenvalues is limited (cf. chapter 8.2.3 \cite{ramsay_functional_2005}). As each curve is infinite-dimensional, there is no upper limit to this number due to the dimensionality. However, the number of curves still imposes an upper limit of $N-1$ non-zero Eigenvalues, where $N$ is the number of curves in the data set.
	These sample analogs naturally lead to the following representation of each sample curve $x_i(t)$.
	\begin{equation}
		x_i(t) = \hat{\mu}(t) + \sum_{j = 1}^{N-1} \hat{\xi}_{i}^{m} \hat{\nu}^{m}(t)
	\end{equation}
	where the $\hat{\xi}_{i}^m$ are derived as 
	\begin{equation}
		\hat{\xi}_i^m(\omega) = \langle x_i - \hat{\mu}, \hat{\nu}^m\rangle = \int_{0}^{1} \left(x_i(s) - \hat{\mu}(s)\right) \hat{\nu}^m(s) \mathrm{d}s
	\end{equation}
	In reality, these calculations are often implemented using basis representations of both the functional principal components $\hat{\nu}^m$ and the observations $x_i(t)$ leading to the following representation. For the sake of clarity, the following equation assumes that the bases used for the expansion of both the observations and the coefficient function are proper bases of $\mathbb{L}^2[0,1]$ and can therefore be used to express the corresponding objects exactly.
	\begin{equation}\label{FPCA_basis_expansion}
		\begin{split}
			\hat{\xi}_{i}^m & = \int_{0}^{1} {\color{red}\left( x_i(s) - \hat{\mu}(s)\right)} {\color{blue}\hat{\nu}^m(s)} \mathrm{d}s
			= \int_{0}^{1} {\color{red}\left(\sum_{j \in \mathcal{I}} a_{i,j} \phi_j(s)\right)} {\color{blue}\left(\sum_{k \in \mathcal{L}} d_{k}^m \psi_{k}(s)\right)} \mathrm{d}s \\
			& = \int_{0}^{1} {\color{red}\left(\sum_{j = 1}^{J} a_{i,j} \phi_j(s) + \delta_i^J(s)\right)} {\color{blue}\left(\sum_{k = 1}^{K} d_{k}^m \psi_{k}(s) + \delta_{\nu}^{m,K}(s)\right)} \mathrm{d}s \\
			& = \sum_{j = 1}^{J} \left[a_{i,j}\sum_{k = 1}^{K} d_{k}^m \int_{0}^{1} \phi_j(s) \psi_{k}(s)\mathrm{d}s \right] +  \sum_{k = 1}^{K} d_{k}^m \int_{0}^{1} \delta_i^J(s) \psi_{k}(s) \mathrm{d}s \\
			& \quad + \sum_{j = 1}^{J} a_{i,j} \int_{0}^{1}\phi_j(s) \delta_{\nu}^{m,K}(s) \mathrm{d}s + \int_{0}^{1} \delta_i^J(s) \delta_{\nu}^{m,K}(s) \mathrm{d}s
		\end{split}
	\end{equation}
	In practice, a typical choice is to use the same basis $\left(\phi_j(t)\right)_{j \in \mathcal{I}}$ and the same truncation parameter $L$ for the basis expansion of the demeaned observations $\left(x_i(t) - \hat{\mu}(t)\right)$ and the functional principal components $\hat{\nu}^m$. This leads to the following simplification of Equation \ref{FPCA_basis_expansion}.
	\begin{equation}\label{score_approx}
		\begin{split}
			\hat{\xi}_{i}^m &= \sum_{j = 1}^{L} \left[a_{i,j}\sum_{k = 1}^{L} d_{k}^m \int_{0}^{1} \phi_j(s) \phi_{k}(s)\mathrm{d}s \right] +  \sum_{k = 1}^{L} d_{k}^m \int_{0}^{1} \delta_i^L(s) \phi_{k}(s) \mathrm{d}s \\
			& \quad + \sum_{j = 1}^{L} a_{i,j} \int_{0}^{1}\phi_j(s) \delta_{\nu}^{m,L}(s) \mathrm{d}s + \int_{0}^{1} \delta_i^L(s) \delta_{\nu}^{m,L}(s) \mathrm{d}s
		\end{split}
	\end{equation}
	And we can define the following objects:
		\begin{align}
			\tilde{\xi}^{m,L}_{i} & := \sum_{j = 1}^{L} \left[a_{i,j}\sum_{k = 1}^{L} d_{k}^m \int_{0}^{1} \phi_j(s) \phi_{k}(s)\mathrm{d}s \right] 
			& \delta_{\xi, i}^L & := \hat{\xi}_{i}^m - \tilde{\xi}^{m,L}_{i} \\
			\tilde{\nu}^{m,L}(t) & := \sum_{k = 1}^{L} d_{k}^m \phi_{k}(t) 
			& \delta_{\nu}^{m, L}(t) & := \hat{\nu}^m(t) - \tilde{\nu}^{m,L}(t)
		\end{align}

@@

	To exploit the regularity of the curves of the spectra, we introduced different basis functions in Section \ref{bases} and demonstrated the importance of the truncation parameter $K$ for the estimation in Section \ref{Scalar_on_function_regression}. For the simulation study, we rely on the introduced estimation strategies with the introduced basis functions and focus on choosing the truncation parameter $L$ as well as the number of FPC's, which is possibly affected by $L$, by ten-fold cross-validation using the prediction mean-squared error. The number of FPC's is in practice chosen by using the lowest number of FPC's that explains a specified proportion of variance (cf. \cite{kokoszka_introduction_2017}), which might not be optimal since FPC's with smaller Eigenvalues may have a more significant influence on the prediction (cf. \cite{Jolliffe_1982}). This might apply to this simulation too since certain Eigenfunctions could correspond to certain chemical combinations and overtones in the absorption bands of the spectrum that could have high predictive power, but explain only little variability in the NIR curves shown in Appendix \ref{NIR}.\\

@@

		 Two different error-terms $\epsilon$ were created by first generating an i.i.d. standard normal error term and then multiplying it by two error variations $\sigma_e $. The error variations represent different signal-to-noise ratios of the responses to test the methods with low and high amounts of noise in the responses. They are created such that the squared multiple correlation coefficient $R^2 = var(\langle X, f\rangle) / (var(\langle X, f\rangle) + \sigma^2_{e})$ is equal to 0.9 and 0.6. The two error terms are then used to generate two sets of responses for $f \in \{f_1(t), f_2(t)\}$	
		\begin{equation}
			\begin{split}
				Y_{1,f} & = \langle NIR_{sim}, f\rangle + Z  \biggl\lbrack\frac{var(\langle NIR_{sim}, f\rangle)}{0.9} - var(\langle NIR_{sim}, f\rangle)\biggr\rbrack \\
				Y_{2,f} & = \langle NIR_{sim}, f\rangle + Z  \biggl\lbrack\frac{var(\langle NIR_{sim}, f\rangle)}{0.6} - var(\langle NIR_{sim}, f\rangle)\biggr\rbrack
			\end{split}
		\end{equation}

@@

	\subsubsection{Basis Expansion Regression}
	The following results originate from the \hyperref[basis_exp_transf]{Estimation using Basis-Representation}, in which we transform the smooth curves to perform scalar-on-function regression. Examining the results, it appears that the cross-validated mean squared prediction error (MSPE) exhibits a convex behavior in the number of basis functions.

	\paragraph{Monomial Basis}
	Due to the increasing collinearity problems for higher numbers of Monomial basis functions, simulations were conducted up until the sixth monomial, which already shows signs of numerical instability. For reasons outlined in \ref{Monomial_basis} {\color{orange} This is not explained anymore in 2.4}, they seem suited for $f_1$, where it shows a better performance than B-splines for this coefficient function. A hypothesis for this could be that $f_1$ is an entire function, which can be well approximated with a power series. In the case of $f_2$, this basis shows the weakest performance out of all basis, for which the plots in Figures \ref{basis_expansion_2_1} and \ref{basis_expansion_2_2} provide evidence. It seems like $\hat{\beta}(t)$ is not changing in the amount of noise and shows exaggerated behavior at the boundaries. This weakness is especially pronounced in the MSPE for $f_2,Y_1$. For $f_1$, the simulation selects 5(3) and for $f_2$ 5(5) Monomial basis functions for the high(low) signal-to-noise ratio. 
	\vspace{-0.2cm}

	\paragraph{B-spline Basis}
	Simulations with B-spline basis functions were possible from 4 to 18 since from 18 onwards, the simulations were running into problems of numerical instability {\color{orange} since the functions are inherently not orthogonal to each other}.
	Function $f_1$ requires 5(4) B-spline basis functions for the high(low) signal-to-noise ratio to obtain the best fit for the B-spline basis and shows the worst performance of the three bases. An explanation might be its overripe behavior at the boundaries and the exaggeration of the peculiarities of $f_1$ (Figures \ref{basis_expansion_1_1} and \ref{basis_expansion_1_2}), which is especially pronounced for the lower signal-to-noise ratio. For $f_2$, 11(6) B-spline basis functions are chosen for the high(low) signal-to-noise ratio. The B-spline basis outperforms the Monomial basis in $f_2$ but comes second to the Fourier basis. While the basis seems to recognize the peculiarities in $f_2,Y_1$, it seems to struggle for the noisy responses in $f_2,Y_2$ (Figures \ref{basis_expansion_2_1} and \ref{basis_expansion_2_2}). With the low signal-to-noise ratio in the responses, the simulation chooses a smaller number of basis functions, which could prevent fitting this scalar noise using $\hat{\beta}(t)$.
	\vspace{-0.2cm}

@@

		Therefore, we will rerun all specifications for the gasoline data set and use the results of the simulation study to improve our understanding of the results. The application is designed analogously to the simulation study: The 60 spectra of the gasoline data set will be used to estimate a coefficient function, predict the reported octane numbers, and evaluate the results via MSPE using 12 fold cross-validation with 5 elements per fold. In total, we conducted 1000 repetitions for each setting.
